\documentclass[10pt,numbers]{sigplanconf}
%\documentclass[10pt,twocolumn]{article}
\usepackage{times}
\usepackage{url}                  % format URLs
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false]{hyperref}   % hyperlinks, including DOIs and URLs in bibliography
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage[pdftex]{graphicx}
\usepackage{tikz}
\usepackage{arydshln}
\usepackage{acmcopyright}


\usepackage{pifont}
\usepackage{amssymb}

%\usepackage{titlesec}
\usepackage{multicol}

\usepackage{pgfplots}

%\usepackage[letterpaper,margin=0.75in,top=1in,bottom=1in]{geometry}

\input{plotsStyle}
\input{macros}

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
%%% Some space-saving stuff added by Idit

\usepackage{etoolbox}
\makeatletter
\patchcmd{\ALG@doentity}{\item[]\nointerlineskip}{}{}{}
\makeatother

\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{2.25ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother

%%% End space saving stuff




\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\onecolumn

\title{\kiwi: A Key-Value Map for Scalable Real-Time Analytics}

\twocolumn

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\authorinfo{{Dmitry Basin\footnotemark[1]}\and {Edward Bortnikov\footnotemark[1]}\and Anastasia Braginsky\footnotemark[1]\and
Guy Golan-Gueta\footnotemark[2]\\ \and   {Eshcar Hillel\footnotemark[1]} \and Idit Keidar\footnotemark[3]\footnotemark[1] \and 
Moshe Sulamy\footnotemark[4]}
{\footnotemark[1]Yahoo Research, Israel \and \footnotemark[2]VMWare Research, Israel \and \footnotemark[3]Technion, Israel
\and \footnotemark[4]Tel Aviv University}
{\{dbasin,ebortnik,anastas,eshcar\}@yahoo-inc.com \and ggolangueta@vmware.com
 \and idish@ee.technion.ac.il \and moshe.sulamy@cs.tau.ac.il}

\date{}

\CopyrightYear{2017}
\setcopyright{acmcopyright}
\conferenceinfo{PPoPP '17,}{February 04-08, 2017, Austin, TX, USA}
%\isbn{978-1-4503-4493-7/17/02}\acmPrice{\$15.00}
%\doi{http://dx.doi.org/10.1145/3018743.3018761}

\maketitle

\begin{abstract}
Modern big data processing platforms employ huge in-memory \emph{key-value} (KV) maps.
Their applications simultaneously drive high-rate data ingestion \emph{and} large-scale analytics.
These two scenarios expect KV-map implementations that scale well with both real-time updates
and large atomic scans triggered by range queries.
%However, today's state-of-the art concurrent KV-maps  fall short of satisfying these requirements.
% -- they either provide only limited or non-atomic scans,  or severely hamper updates when scans are ongoing.

We present {\kiwi}, the first atomic KV-map to efficiently support simultaneous  large scans and real-time access.
The  key to achieving this is treating scans as first class citizens, and organizing the data structure around them.
%whereas most existing concurrent KV-maps do not provide
%atomic scans, and some others add them to existing maps without rethinking the design anew.
\kiwi\ provides {wait-free} scans,
%which is important for avoiding livelock and wasted work,
whereas its put operations are lightweight and lock-free.
It optimizes memory management jointly with  data structure access.
%We prove \kiwi's correctness and also implement it
We implement \kiwi\
and compare it to state-of-the-art solutions.
%our evaluation shows that \kiwi\ is significantly faster than the state-of-the-art.
Compared to other  KV-maps providing atomic scans,
%in workloads with long scans and concurrent writes,
\kiwi\ performs either long scans or concurrent puts an order of magnitude faster. Its   scans are twice as fast as  \emph{non-atomic} ones implemented
via iterators in the Java skiplist.


\inote{
A number of novel aspects of the {\kiwi} algorithm make it particularly well-suited for today's data processing systems.
First, it provides wait-free scans, which is important for avoiding livelock and wasted work, given that snapshot scans are often long.
On the other hand, given that updates are short, \kiwi\ opts to make them lock-free rather than wait-free, since restarts due to conflicts
are practically rare and waste little work.  Second, {\kiwi} uses multi-versioning, but keeps old data versions selectively,
only as needed for ongoing scans, and furthermore keeps the updates light-weight by deferring version management to scans.
Third, the {\kiwi} algorithm optimizes memory management and reclamation jointly with the optimization of data structure access.
Finally, {\kiwi} implements a balanced data structure, which allows operations to run orders of magnitude faster than
unbalanced ones in case update order is not random.
}

\end{abstract}
%\bigskip
%\bigskip
%\centerline{Corresponding author: Anastasia Braginsky aGInastas@yahoo-inc.com}
%\noindent
%{\bf
%\centerline{The paper is eligible for the best student paper award.}
%\centerline{Moshe Sulamy is a full-time PhD student in Tel Aviv University\footnote{
%This work was done in part while interning with Yahoo Labs, Haifa. }.}
%}

%\end{titlepage}

\renewcommand{\thefootnote}{\arabic{footnote}}

\input{intro}

\section{Related Work}
\label{sec:related}
\input{related}

\input{algorithm1}
\input{proof1}
\input{eval}

\section{Discussion}
\label{sec:disc}

We presented {\kiwi}, a KV-map tailored for real-time analytics applications. \kiwi\ is the first
concurrent KV-map to support high-performance atomic scans simultaneously with real-time updates of the data.
In contrast to traditional approaches, {\kiwi} shifts the synchronization
overhead from puts to scans, and offers lock-free puts and wait-free gets and scans.
We demonstrated {\kiwi}'s significant performance gains over state-of-the-art KV-map
implementations that support atomic scans. 

\inote{
Future work could apply {\kiwi} as the in-memory part of a NoSQL KV-store, e.g., HBase~\cite{ApacheHBase}.
Prior results~\cite{GolanGueta2015} highlighted that the overall  performance of such systems can be boosted by
improving  concurrency at the memory part.%in-memory KV-store.
}

\bibliography{ref}
%\bibliographystyle{abbrv}
\bibliographystyle{abbrvnat}
%

\end{document}




